Data Analysis Methods for the Detection of the Epoch of Reionization

### 2017-06-29 PAPER.tex
finished rough draft of power spectrum estimation section

To form power spectrum quantities using our two fringe-rate filtered, LST-binned datasets, we use OQE methods as described by \cc{??}. A summary of the method is as follows. We begin with our data vector $\textbf{x}$, which contains our visibilities that have the shape ($times$, $frequencies$) for each baseline and data group (`even' or `odd'). As part of our bootstrapping routine (and to speed things up), we divide up our total number of baselines into $5$ random, equal-sized groups, where no baselines are repeated within or between groups. For the \citet{ali_et_al2015} method, each group is then sampled with replacement to create a new group, of the same size, that can have repeated baselines inside it. We discover that in doing so, we are sacrificing some of our sensitivity since this results in there being 3-4 repeated baselines per group. In order to maximize our sensitivity but still apply random sampling for use in error estimation, we instead form new groups using all independent baselines except the very last one. For example, if we have $10$ baselines in a group, we use the first $9$, and then fill the last slot randomly out of the $10$. We do this for all $5$ groups. This is still a valid means of bootstrapping because there are many more possibilities of baseline groupings than the number of bootstraps we run ($20$).

@@

Notice that there are two quantities of $\textbf{C}^{-1}\textbf{x}$ in the expression for $\hat{q}$, representing two copies of inverse covariance weighted data. We perform all possible cross-multiplications of this quantity, except between the same datasets (`even' with `even', for example) and same groups (`baseline group 1' with `baseline group 1', for example). We avoid these computations to prevent auto power shared between the two quantities from introducing biases.

@@

where $\hat{\textbf{p}}$ is the normalized estimate of the true power spectrum $\textbf{p}$. $\textbf{M}$ is related to the Fisher matrix $\textbf{F}$ by:

\begin{eqnarray}
\textbf{F}_{\alpha\beta} &=& \frac{1}{2}tr(\textbf{C}^{-1}\textbf{Q}_{\alpha}\textbf{C}^{-1}\textbf{Q}_{\beta}) \\
\textbf{W} &=& \textbf{MF}
\end{eqnarray}

@@

Our final power spectrum quantities are 1-dimensional functions of $k$. We compute $20$ bootstraps of these quantities, where each bootstrap creates baseline groups randomly.

\cc{Explain new PS channels, like noise and eor, and how they're created}

@@

\cc{Add info on how we calculate effective inttime and adjust beam factor because of FRF}

>>>>

To form power spectrum quantities using our two fringe-rate filtered, LST-binned datasets, we use OQE methods as described by \cc{??}. A summary of the method is as follows.

We begin with our data vector $\textbf{x}$, which contains our visibilities that have the shape ($times$, $frequencies$) for each baseline and data group (`even' or `odd'). As part of our bootstrapping routine (and to speed things up), we divide up our total number of baselines into $5$ random, equal-sized groups, where no baselines are repeated within or between groups. For the \citet{ali_et_al2015} method, each group is then sampled with replacement to create a new group, of the same size, that can have repeated baselines inside it. We discover that in doing so, we are sacrificing some of our sensitivity since this results in there being 3-4 repeated baselines per group. In order to maximize our sensitivity but still apply random sampling for use in error estimation, we instead form new groups using all independent baselines except the very last one. For example, if we have $10$ baselines in a group, we use the first $9$ to guarantee at least $9$ independent measurements, and then fill the last slot randomly out of the $10$. We do this for all $5$ groups. This is still a valid means of bootstrapping because there are many more possibilities of baseline groupings than the number of bootstraps we run for this analysis ($nboots = 20$).

@@

Notice that there are two quantities of $\textbf{C}^{-1}\textbf{x}$ in the expression for $\hat{q}$, representing two copies of inverse covariance weighted data. We perform all possible cross-multiplications of this quantity, except between the same datasets (`even' with `even', for example) and same groups (`baseline group 1' with `baseline group 1', for example). We avoid these computations to prevent the introduction of biases from shared auto power.

@@

where $\hat{\textbf{p}}$ is the normalized estimate of the true power spectrum $\textbf{p}$. We compute $\textbf{M}$ using the Fisher matrix $\textbf{F}$, defined as:

\begin{eqnarray}
\textbf{F}_{\alpha\beta} &=& \frac{1}{2}tr(\textbf{C}^{-1}\textbf{Q}_{\alpha}\textbf{C}^{-1}\textbf{Q}_{\beta}). \\
\end{eqnarray}

We have a choice for $\textbf{M}$, and we choose to compute it by taking the Cholesky decomposition of $\textbf{F}$. Namely, we decompose the Fisher matrix such that $\textbf{F} = \textbf{L}\textbf{L}^{\dagger}$, where $\textbf{L}$ is a lower triangular matrix. Next, we construct $\textbf{M} = \textbf{D}\textbf{L}^{-1}$, where \textbf{D} is a diagonal matrix. Hence, our window function $\textbf{W} = \textbf{MF}$ becomes $\textbf{W} = \textbf{D}\textbf{L}^{\dagger}$ and is an upper triangular matrix. This window function was constructed in a way to prevent the leakage of foreground power from low $k$ to high $k$ modes. Specifically, we order the elements in $\textbf{F}$ in such a way so that power can leak from high $k$ modes to low $k$ modes, but not vice versa. Since most foreground power shows up at low $k$'s, this method ensures a window function that retains clean, noise-dominated measurements.

@@

In practice, we end up with a 2-dimensional power spectrum quantity, which is a function of both time and $k$. We compute $20$ bootstraps of these quantities, where each bootstrap creates baseline groups randomly.

In \citet{ali_et_al2015}, a second round of bootstrapping occurs over the bootstrap and time axes simultaneously. Random values are sampled with replacement along both axes, drawing as many values as there are number of bootstraps and times. Final power spectrum limits are then computed by taking the mean and standard deviation over this second bootstrap axis.

However, we have found that this method greatly underestimates power spectrum errors, especially for fringe-rate filtered data. This can be explained by the fact that fringe-rate filtered data has a dramatically reduced number of independent modes. Hence, drawing $100$ samples out of a length-$100$ dataset that only has $5$ independent modes in it, for example, results in a narrower distribution of values that leads to a false error estimation.

To avoid this issue, we instead take a simple average along the time axis. Our final power spectrum limits are computed by taking the mean and standard deviation over our single bootstrap axis.

In summary, our power spectrum pipeline begins with data with four axes (datasets, baselines, times, frequencies), bootstraps over baselines, uses OQE formalism, and produces $1$-dimensional power spectra for both $\textbf{P(k)}$ and the folded version $\Delta^{2}\textbf{(k)}$, defined as:

\begin{equation}
\Delta^{\textbf{2}}\textbf{(k)} = \frac{k^{3}}{2\pi^{2}}\hat{\textbf{p}}\textbf{(k)}
\end{equation}

One critical new component of our power spectrum pipeline is that we have many different power spectrum channels that simultaneously get processed at the same time and are especially useful for signal loss (Section \ref{sec:Sigloss}) and sensitivity (Section \ref{sec:Sense}) verification. Two important channels, in addition to our PAPER-64 data, are a noise dataset and an EoR dataset (in addition to these, we do different combinations of all three). We will briefly describe each.

We create random noise with a scale determined by the sensitivity of our instrument and parameters of our dataset. We calculate our system temperature for our frequencies of interest as:

\begin{equation}
T_{sys} = 180\Big(\frac{\nu}{0.18}\Big)^{-2.55} + T_{rcvr}
\end{equation}

where $\nu$ are frequencies in GHz. We use a receiver temperature of $200$ K, yielding $T_{sys} = 487$ at $150$ MHz. This is lower than in \citet{ali_et_al2015} because \cc{why?}. We convert this temperature into a variance statistic using:

\begin{equation}
T_{rms} = \frac{T_{sys}}{\sqrt{\Delta\nu \Delta t N_{days} N_{pols}}}
\end{equation}

where $\Delta\nu$ is channel spacing, $\Delta t$ is integration time, $N_{days}$ is the number of data samples for a particular time and frequency that went into our LST binned set, and $N_{pols}$ is the number of polarizations ($2$ for Stokes I). Our simulated noise has the same shape as our data, and we fringe-rate filter it in the same way to best mimic the real noise floor of our data.

A second additional important channel in our pipeline is a simulated EoR signal. We again simulate random noise (with a default scale of $1$) with the same shape as our data, and we fringe-rate filter this signal twice. The first filter transforms the unattached noise into a signal that's attached to the sky (what our instrument observes). The second filter represents the fringe-rate filtering step in our data analysis pipeline. As described in Section \ref{sec:Sigloss}, we create an EoR signal at various amplitude levels.

Our multi-channel power spectrum pipeline has been essential in helping us understand the nuances in our pipeline. Because there are many different components and countless subtle effects affecting our final limits, it has been imperative to carry all our channels through the analysis to validate each step.

@@

\cc{Add info on sensitivity equation, how we calculate effective inttime and adjust beam factor because of FRF, etc.}
